---
title: "Cod and Crab CCM"
author: "Owen Liu"
date: "January 7, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(rEDM)

# figure theme
plot_theme <-   theme_minimal()+
  theme(text=element_text(family="sans",size=12,color="black"),
        legend.text = element_text(size=14),
        axis.title=element_text(family="sans",size=14,color="black"),
        axis.text=element_text(family="sans",size=8,color="black"),
        panel.grid.major = element_line(color="gray50",linetype=3))
theme_set(plot_theme)
```

# Pacific Cod and Snow Crab: Evidence of Dynamic Causation?

Here we apply convergent cross mapping to time series data of pacific cod and snow crab to test if there is evidence that they drive one another's dynamics, specifically:

*Is there evidence that Pacific Cod abundance drives Snow Crab recruitment? Conversely, does Snow Crab recruitment drive Pacific Cod abundance?*

## Import data

```{r}
load('data/cod_dat_clean.Rdata')
load('data/longform_opilio.Rdata')
```

For now, the only cod data we have is total weight and total numbers, by station, by year. We do not have sex, size, or age composition.

For snow crab, we have sex, size, and maturity. However, for the purposes of this first analysis, we will use time series of recruits.

The data are cleaned from the annual summer bottom trawl survey in the Eastern Bering Sea (EBS).

Finally, for simplicity's sake, for now we will define "recruits" as snow crab <38 mm CW

```{r}
opi_recruits <- opi_dat %>% 
  filter(Size<38,units=='number')

# normalized cod time series
codnorm <- cod_dat_clean %>% 
  select(station,Year,dens_weight) %>%
  arrange(station,Year) %>% 
  group_by(station) %>% 
  # fill in explicit missing years
  complete(Year=full_seq(Year,1)) %>%
  ungroup() %>% 
  mutate(normdens=(dens_weight-mean(dens_weight,na.rm=T))/sd(dens_weight,na.rm=T))

codsegs <- codnorm %>% 
  ungroup() %>% 
  mutate(n=row_number()) %>% 
  group_by(station) %>% 
  summarise(first=first(n),last=last(n)) %>% 
  select(-station) %>% ungroup()
```

## Simplex for cod
```{r}
codsimp <- simplex(codnorm$normdens,lib=as.matrix(codsegs))
```


