---
title: "Environmental Variability Drives Predation Risk: A Bering Sea Example"
author: "Liu, O.R., Szuwalski, C.S., Gaines, S.D."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: bookdown::pdf_document2
editor_options:
  chunk_output_type: inline
fontsize: 11pt
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
- \renewcommand\linenumberfont{\normalfont\bfseries\small}
csl: nature.csl
bibliography: library.bib
---

```{r global options, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning = FALSE)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abstract

# Introduction

Managed species are important
There is a push for ecosystem-based fisheries management
There are two major things we are not good at characterizing- environmental variability and species interactions
We know that interactions can play a large role in population dynamics
Pacific cod and snow crab are two of the largest and most lucrative fisheries in the eastern Bering Sea
  Both have declined (show some stats); cod collapses elsewhere
  We know that cod eat crab
  We know that both are influenced by temperature and ontogenetic dynamics
  What we don't know is how these influences interact
We use spatiotemporal models and spatially-resolved diet data to characterize the relationships between cod and snow crab dynamics and spatial distributions
Our approach helps to sort out the complex interactions between dynamic species distributions, the environment, and interactions

The goal of ecosystem-based fisheries management (EBFM) is to promote sustainable marine ecosystems and fisheries through consideration of interactions between humans and both biotic and abiotic elements of the natural environment. Although EBFM has been actively pursued for decades[@Pikitch2004;@Fogarty2014] implementation has been a difficult goal to achieve[@Pitcher2009]. One of the reasons for the slow adoption of EBFM is the technical difficulty inherent in trying to simultaneously account for the influences of species interactions and varying environmental conditions on species' abundances and distributions across space[@Orio2019].

Some fishery collapses or lack of recoveries have been attributed to a failure to account for species interactions[@Hollowed2000;@Swain2012]. Importantly, declining to include species interactions like predation into models of population dynamics may lead to an underestimation of uncertainty and perhaps to overly optimistic or risky management advice[@Hollowed2000a]. At the same time, environmental drivers act on these interacting species, determining phenology, patterns of species movements, recruitment of new individuals into populations, and by extension, overall productivity[@Pagel2012;@Szuwalski2015]. When environmental change drives greater spatial overlap between predators and prey, interactions will intensify and the relative influence of predation will strengthen[@Kempf2010]. In this way, environmental variability can determine patterns of species interactions.

In fisheries that have declined from historical production levels, there are often attempts to attribute causation to either species interactions or environmental change in addition to fishing pressure[@Frank2016;@Swain2019]. One such fishery is the snow crab (*Chionoecetes opilio*) fishery in the eastern Bering Sea (EBS). The fishery for snow crab is one of the most valuable crustacean fisheries in the world, but has experienced marked fluctuations in stock size over the course of its exploitation history[@Parada2010]. Production in the snow crab fishery peaked in the late 1980s, but landings declined by more than an order of magnitude from the early 1990s to the early 2000s. However, the snow crab stock has recently been rebuilt after a string of strong year classes since 2010[@NPFMC2018].

A number of hypotheses have been suggested to explain the fluctuations in the snow crab population in the EBS, including oceanographic and climate forcing[@Zheng2006;@Parada2010;@Orensanz2004]. Juvenile snow crab settlement has been correlated to the annual spatial extent of the "cold pool" across the middle portion of the shelf, an area with near-bottom temperatures < 2$^{\circ}$C which is created by severe stratification of the water column as ice melts across a large portion of the Bering Sea in the spring[@Wyllie-Echeverria1998;@Ernst2005]. After warmer winters, the cold pool extent is reduced to the northwest, while after colder winters it can extend across a wide area of the shelf to the southeast(FIGURE INTRO MAP). Moreover, after warm years as the snow crab population contracts to the northwest, the stock can have trouble recolonizing the southern portion of the shelf. Orensanz et al. (2004)[@Orensanz2004] postulated that this is due to the fact that prevailing currents in the region flow in a northwesterly direction, preventing larvae from being advected "upstream" to other portions of the shelf, even after environmental conditions have returned to normal[@Ernst2012].

Predation by fish predators on the juvenile snow crabs may be another important driver of their spatial populations dynamics[@Livingston1989;@Orensanz2004;@Burgos2013]. Previous research suggests that the southern boundary of snow crab distribution in the Bering Sea is enforced via predation by Pacific cod (*Gadus macrocephalus*). Pacific cod in some years consume enormous amounts of juvenile snow crab, just as Atlantic cod predate on snow crab in eastern Canada[@Livingston1989;@Chabot2008]. This effect of cod predation on snow crab mortality is size depedent for both species[@Burgos2013]; that is, the size of snow crab consumed by cod is dependent on both crab size and cod size. It has been suggested that this predation effect can interact with and exacerbate the cold pool effect on juvenile snow crab settlement[@Orensanz2004]. Pacific cod distribution also responds to environmental variability. In warmer years the centroid of adult cod distribution moves to the northwest[@Zheng2006], although juvenile cod distribution is relatively stable across the EBS[@Hurst2012]. When snow crab distribution retracts to the northwest in warmer years, the expanded cod distribution may be an additional factor preventing the reexpansion of the stock towards the southeast.

In this study, we investigate the spatial distributions of snow crab and Pacific cod. Using spatial dynamic factor analysis (SDFA) applied to annual trawl survey data, we investigate the consistent spatial and temporal patterns in cod and crab distributions by size. We examine the time period from the early 1980s to the present (2016), a time period characterized by large fluctuations in both crab and cod abundance. The hypothesis is that cod and crab will both respond to fluctuations in patterns relating to environmental drivers, specifically bottom temperature and depth. In addition, after constructing a spatio-temporal model of crab and cod distributions, we test whether cod biomass is correlated with lagged crab biomass in the EBS, which would be suggestive of a significant predation effect.

# Data and Methods

## Trawl survey data

To investigate the spatial patterns in Pacific cod and snow crab distributions as they vary over time, we use a joint dynamic species distribution model(JDSDM)[@Thorson2016]. JDSDMs estimate "factors" representing latent spatial and spatio-temporal patterns in observed data. We fit the model to fishery-independent bottom trawl survey data from the Gulf of Alaska. The survey, conducted each summer, enumerates all species caught in each of approximately 375 tows across a 20 by 20 nautical mile grid, providing an annual census of the Bering Sea demersal fish and invertebrate communities (full survey details can be found at https://www.fisheries.noaa.gov/resource/document/groundfish-bottom-trawl-survey-protocols). Along with the number and weight of species caught, the survey also records near-bottom temperature, depth, and area swept by the trawl. For both cod and crab, all individuals are sexed and measured by fork length (FL) for cod or carapace width (CW) for crabs. Additionally, the survey denotes crab maturity stage, enabling us to distinguish between immature and mature crabs.

From the survey data, we extract all observations of snow crab and cod, then sort them into size bins. For snow crab, we defined two classes: immature and mature females. We defined immature crabs as crabs of both sexes that were smaller than 58 mm CW and sexually immature. The 58 mm cutoff is based on a diet study that found that 95% of all crabs found in Pacific cod stomachs are smaller than 58mm. We define mature female crabs as any mature females, regardless of egg-carrying status or size. Size at maturity for snow crabs varies with latitude and year, although in most regions and years, primipara (first-time female spawners) are larger than 50 mm CW[@Ernst2012].

For Pacific cod, we enumerate total abundance at each survey station for cod within three size classes. As with snow crab, our size classes were based on previous studies of cod predation, which found that cod containing snow crabs in their stomachs were bewtween 200 and 800 mm FL[@Livingston1989;@Burgos2013]. We define small cod as those <200 mm FL, medium cod as between 200 and 800 mm, and large cod as larger than 800 mm. Therefore, with our definitions, we would expect a predation interaction between medium cod and immature snow crab. Size frequency distributions for all classes used in the model are shown in FIGURE XXX. In the following, we refer to the two crab and three cod categories in the data as "classes".

## Joint dynamic species distribution model

We implement the JDSDM through the publically available Vector Autoregressive Spatiotemporal R package `VAST`. VAST uses a delta-generalised linear mixed modeling method, and takes into account spatio-temporal correlations among species (or size classes of one species, as in our model). A full description of the modeling framework can be found elsewhere[@Thorson2016;@Thorson2019], but we describe the key equations here.

The JDSDM models response variables as a multivariate process where the predicted abundance of a class at a location is represented as the combination of two linear predictors: encounter rate (i.e., presence/absence of a class at a given location), and positive catch rate (i.e., the prediction of abundance given the presence of a class in a location). VAST estimates fixed and random effects for the two linear predictors that are each a function of an intercept, a spatial effect, a temporal effect, and any density covariates. In our model, bottom temperature and depth were used as density covariates. More specifically, the first linear predictor represents encounter rate for sample $i$ as,

\begin{equation}
  p_1(i)=\beta_1(c_i,t_i)+\sum_{f=1}^{n_{\omega1}}L_{\omega1}(c_i,f)\omega_1(s_i,f)+\sum_{f=1}^{n_{\epsilon1}}L_{\epsilon}(c_i,f)\epsilon_1(s_i,f,t_i)+\sum_{p=1}^{n_p}\gamma_1(c_i,t_i,p)X(s_i,t_i,p)
  (\#eq:linpred)
\end{equation}

where $p_i$ is the linear predictor for encounter rate, $\beta_1(c_i,t_i)$ is the intercept term in year $t$ for class $c$, the terms starting with $L_\omega$ and $L_\epsilon$ are the spatial and spatio-temporal factor models, respectively, and $X$ is the matrix of density covariates defined for each location $s$, time $t$ and covariate $p$ that have linear effects $\gamma$ on the predictor. In the factor model terms, $L\omega$ and $L_\epsilon$ are loading matrices relating the classes to each of $f$ factors, and $\omega(s,f)$ and $\epsilon(s,f,t)$ are vectors of random effects representing latent spatial and spatio-temporal variation at each location. VAST models average spatial variation $\omega$ as being constant across years, while spatio-temporal variation $\epsilon$ varies among years. The second linear predictor, representing positive density, has an analogous formulation.

We use a log-link to transform both linear predictors to predict the observed data (see Thorson 2019 for all model equations):

\begin{equation}
  r_1(i) = 1-exp(-a_ixexp(p_1(i)))
  r_2(i) = \frac{a_ixexp(p_1(i))}{r_1(i)xexp(p_2(i)))
  (\#eq:transpred)
\end{equation}

where $a_i$ is the area associated with sample $i$ in the survey. Positive abundance is modeled with a gamma distribution.

An important feature of VAST is that it accounts for spatial autocorrelation, or that fact that values nearby in space tend to be more similar than locations further away. Each spatial ($\omega$) and spatio-temporal ($\epsilon$) factor in both linear predictors are represented as Gaussian random fields, where the covariance between nearby locations is approximated with a Matern correlation function[@Lindgren2011] through the R-INLA software[@Illian2012]. The correlation function follows geometric anisotropy, where the rate of decline in correlation between locations decline can be different in different directions (i.e., correlation can decline more rapidly in the east-west than the north-south direction).

Parameter estimation for VAST is accomplished through Laplace approximation of the marginal likelihood for fixed effects while integrating across the random effects using the Template Model Builder software[@Kristensen2015]. The model requires the estimation of 87 fixed effects and 25752 random effects. For this study, we are primarily interested in the spatial and spatio-temporal factors, and how the five classes in our study relate to those factors, represented by the factor loadings matrices. We estimated three spatial and three spatio-temporal factors for both linear predictors. As derived quantities, VAST can also calculate each class' spatial abundance-weighted center of gravity and effective area occupied within the EBS, as well as an overall (non-spatial) index of abundance. We use these derived quantities to further explore the spatial and temporal dynamics of snow crab and cod.

# Results

We estimated three factors for spatial and spatio-temporal variation in the five classes (two snow crab and three Pacific cod classes, Figure XXX). To visualize the factors, we estimated all spatial values at 100 distinct spatial locations, analytically chosen to minimize the distance between samples and knots.

Description of factor trends and relationship to environmental drivers with correlations.
General time series, spatial factors, and range indices
Correlations in encounter rates and predicted densities
Lagged correlations in space for each knot


# Discussion

# References
